---
title: "Getting Started with hclust1d"
author: "Szymon Nowakowski"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# This document

The purpose of this vignette is to introduce readers to `hclust1d` package, 
and bring them up to speed by providing a simple
use case example.


# `hclust1d` package

The name of `hclust1d` package stands for Hierarchical CLUSTering for 1D. The package contains a suit of algorithms for univariate agglomerative hierarchical clustering (with a comprehensive list of choices of a linkage function, matching all possible choices in `stats::hclust`).

Hierarchical clustering in the 1D setting works in $\mathcal{O}(n\log n)$ time. 

# OK, so let's get started

To load the package into memory execute the following line:

```{r setup}
library(hclust1d)
```

We will work with random normally distributed data points in this vignette.

Working with `hclust1d` is very simple an requires only passing the data points vector and
optionally a linkage method to `hclust1d` function (complete linkage is used as a default, if the linkage method is not provided). The simplest example of a complete linkage clustering:

```{r}
dendrogram <- hclust1d(rnorm(100))
```

The `hclust1d` function returns an object of the same type that is returned from `stats::hclust` (a list object with S3 class \code{"hclust"}, to be specific).
This makes it straightforward to further work with the clustering results:

- we can plot it:

```{r}
plot(dendrogram)
```

- we can print it:

```{r}
print(dendrogram)
```

- we can derive a summary of a result, too:

```{r}
summary(dendrogram)
```
