---
title: "A guide on replacing `stats::hclust` with `hclust1d`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A guide on replacing `stats::hclust` with `hclust1d`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# This document

The purpose of this vignette is to provide guidelines on replacing `stats::hclust` calls with
`hclust1d` calls for univariate (1D) data, in a plug-and-play manner, i.e. without changing any surrounding code (or little of the surrounding code, as an option to the programmer).

To enable use of `hclust1d` you need to include this line in your script or markdown notebook:
```{r setup, message = FALSE}
library(hclust1d)
```

In case of packages you need to import `hclust1d` in your `DESCRIPTION` file.

# Why bother?

The main reason a programmer might want to replace `stats:hclust` calls with
`hclust1d` in case of clustering 1D points is that the computational complexity of `hclust1d` is $\mathcal{O}(n\log n)$, while it is at least quadratic (and $\mathcal{O}(n^2\log n)$ for case of general linkage) for multidimensional algorithms.

The better algorithmic time complexity (compared to multidimensional hierarchical clustering) paired with its efficient `C++` implementation make `hclust1d` very fast. The computational time beats `stats::hclust` on all sizes of data and is *en par* with `fastcluster::hclust` with small data sizes. However, it is of orders of magnitude faster than both multivariate clustering routines on larger data sizes.

# Prerequisities

### Compatibility of `hclust1d` with `stats::hclust`

Maintaining compatibility of `hclust` with `stats::hclust`was high on a list of design priorities for `hclust1d`. 

1. All linkage functions of `stats::hclust` are supported in `hclust1d`, too. 
2. Input to `stats::hclust` should be `dist` S3 class structure as produced by `stats::dist` function and the same input is accepted in `hclust1d` (with a `distance` argument explicitly set to `TRUE`).
3. There are three atypical linkages in `stats::hclust`. Namely, `stats::hclust` requires that the **squared** `dist` structure is provided for `ward.D`, `centroid` and `median` linkage functions. This is implicit. The same input is accepted in `hclust1d` (with `distance` and `squared` arguments explicitly set to `TRUE`).
4. The object returned from `hclust1d` call is the same S3 class as the result of `stats::hclust`, namely `hclust` S3 class.
5. The heights returned from `hclust1d` are calculated the same, as in `stats::hclust`:

   a. **squared** respective distances are returned for `ward.D`, `centroid` and `median` linkage functions,
   b. **unsquared** respective distances are returned for the remaining linkage functions.

### Choosing a linkage function in `hclust1d`

The list of all linkage functions supported in `hclust1d` is available by calling:
```{r supported_methods}
supported_methods()
```

The in-depth description of the linkage functions in `hclust1d`, together with the inter-cluster distance metric definition used in case of each linkage function (and returned as the merging height) can be found in our [getting started vignette](https://cran.r-project.org/package=hclust1d/vignettes/getting-started.html).

The choice of a linkage function is the same in `hclust1d` as in `stats::hclust`, i.e. by specifying a `method` argument and passing the name of a linkage function into `hclust1d` as a character string.

To provide an example, the following two calls execute `average` linkage hierarchical clustering on distances computed for a set of 1D points, by passing `method = "average"` argument to relevant calls:
```{r average}
points <- rnorm(10)

res <- stats::hclust(stats::dist(points), method = "average")
res <- hclust1d(stats::dist(points), method = "average", distance = TRUE)
```

### Choosing a distance metric in `hclust1d`

The user of `stats::hclust` and of `hclust1d` can select a number of distance metrics when building distance-based input with `stats::dist`, by selecting an appropriate name of a metric and passing it as a `method` argument to `stats::dist` as a character string. Not all of them are supported in `hclust1d`. The list of distance metrics supported in `hclust1d` is available by calling:
```{r dist_supported_methods}
supported_dist.methods()
```

The trick here is that for 1D points `euclidean`, `maximum`, `manhattan` and `minkowski` distances are equivalent. 

To provide an example, the following two calls execute `average` linkage hierarchical clustering on distances computed by `minkowski` $L_3$ norm for a set of 1D points, by passing `method = "minkowski"` and `p = 3` arguments to relevant `stats::dist` calls:
```{r minkowski}
points <- rnorm(10)

res <- stats::hclust(stats::dist(points, method = "minkowski", p=3), method="average")
res <- hclust1d(stats::dist(points, method = "minkowski", p=3), method="average", distance = TRUE)
```


### Exceptions

We don't support `members` argument in `hclust1d`.

# Replacing `stats::hclust` in case of `ward.D`, `centroid` or `median` linkage functions

**This section DOES NOT apply to *ward.D2* linkage function, despite the similarity in its name**.

As can be seen from the above sections, to replace a `stats::hclust` call with `hclust1d` for 1D data, one needs to replace any call to
```{r eval=FALSE}
res <- stats::hclust(squared_d, method = linkage_function_name, members = NULL)
```
by a call to
```{r eval=FALSE}
res <- hclust1d(squared_d, method = linkage_function_name, distance = TRUE, square = TRUE)
```

Somewhere in the code above this line, `squared_d` has been computed by a call to
`stats::dist` from a vector of 1D points and subsequently squaring the `stats::dist` result.

Somewhere below in the code `res` gets analyzed, but it is OK, because the results of both calls are compatible.

### Optional changes to the surrounding code

- If the programmer has access to the original `stats::dist` result (let's denote this variable `d`), the computation of `squared_d` can be removed (provided it is not used for other purpose) and a call to
  ```{r eval=FALSE}
  res <- stats::hclust(squared_d, method = linkage_function_name, members = NULL)
  ```
  can be replaced
  by a call to
  ```{r eval=FALSE}
  res <- hclust1d(d, method = linkage_function_name, distance = TRUE)
  ```

- If the programmer has access to the original points (let's denote this variable `points`), the computation of `squared_d` and of `d` can be removed altogether  (provided they are not used for other purpose) and a call to
  ```{r eval=FALSE}
  res <- stats::hclust(squared_d, method = linkage_function_name, members = NULL)
  ```
  can be replaced
  by a call to
  ```{r eval=FALSE}
  res <- hclust1d(points, method = linkage_function_name)
  ```

# Replacing `stats::hclust` in case of all other linkage functions, besides `ward.D`, `centroid` or `median`

**This section applies to, among others, *ward.D2* linkage function**.

As can be seen from the above sections, to replace a `stats::hclust` call with `hclust1d` for 1D data, one needs to replace any call to
```{r eval=FALSE}
res <- stats::hclust(d, method = linkage_function_name, members = NULL)
```
by a call to
```{r eval=FALSE}
res <- hclust1d(d, method = linkage_function_name, distance = TRUE)
```

Somewhere in the code above this line, `d` has been computed by a call to
`stats::dist` from a vector of 1D points.

Somewhere below in the code `res` gets analyzed, but it is OK, because the results of both calls are compatible.

### Optional changes to the surrounding code

If the programmer has access to the original points (let's denote this variable `points`), the computation of `d` can be removed (provided it is not used for other purpose) and a call to
```{r eval=FALSE}
res <- stats::hclust(d, method = linkage_function_name, members = NULL)
```
can be replaced
by a call to
```{r eval=FALSE}
res <- hclust1d(points, method = linkage_function_name)
```
